### 2017-11-09 latex/mst_lukkarila_juri_2017.tex
edit

\textit{Automatic speech recognition} (ASR) is the science and technology of transcribing spoken language into written words automatically using computers \cite{yu2014automatic, huang2001spoken}. The history of speech recognition research goes as far back as the 1950s, when the first steps were taken at the renowned Bell Laboratories \cite{gales2008application}. For a long time, working applications were comprised mostly of simple voice user interfaces and dictation systems with limited, application-specific vocabularies \cite{yu2014automatic, gales2008application}. The holy grail of speech recognition technology has arguably been real-time, speaker independent recognition of unlimited vocabularies, of which everyday conversational speech is a good example. In recent years, significant advances have been made especially in this area, referred to in ASR research as \textit{large vocabulary continuous speech recognition} (LVCSR) \cite{yu2014automatic, keronen2014approaching}. These advances have been enabled in large part thanks to the fast development and adoption of new machine learning techniques \cite{yu2014automatic, hinton2012deep}. In particular, \textit{deep neural networks} (DNN) with multiple hidden layers (hence the word "deep") have proven to be of great value for ASR systems \cite{yu2014automatic, hinton2012deep}. Machine learning itself has become widespread and practical owing to the advances in computer hardware and computational resources available, especially through the popularization of the so called \textit{general-purpose computing on graphics processing units} (GPGPU) technology for massively parallel computation, that has enabled efficient training of neural networks with relatively inexpensive and widely available computing hardware \cite{yu2014automatic, hinton2012deep}. \\\\
As a consequence, automatic speech recognition has gained a great deal of popularity during the last few years, and is starting to be widely used in practice in the commercial landscape and everyday technical applications, from smartphones and tablets to personal computers \cite{yu2014automatic}. Today, companies actively promoting their own speech recognition systems to average consumers include influential giants like Amazon, Apple, Google and Microsoft. Services based on speech recognition technology encompasses voice-controlled personal assistants, dictation, automatic captioning, voice-based search of audio and video content, and a wide variety of voice user interfaces \cite{yu2014automatic}. \\\\
While there are still some specific, and in many cases, quite substantial challenges for automatic speech recognition, in general it has become accurate and reliable enough to be used in many practical applications requiring continuous recognition of large vocabularies \cite{yu2014automatic, keronen2014approaching, mcgraw2016personalized}. This is demonstrated well by the speech recognition based services of the previously mentioned large technology companies, such as Apple's Siri voice assistant. At the same time, mobile smart devices (phones, tablets and laptops) and  wireless internet access have become ubiquitous in all developed countries of the world, offering a convenient platform for utilizing speech recognition technology in practically every conceivable place and situation \cite{yu2014automatic, mcgraw2016personalized}.

@@

In this work, the primary interest in automatic speech recognition is in using it in practice, instead of developing or improving upon some aspect of it. Therefore, this section focuses on providing a high-level, short overview of ASR systems. Likewise, detailed description of machine learning techniques and training methods fall outside this thesis. The focus in this thesis is on modern large vocabulary continuous speech recognition. \\\\ An automatic speech recognition system takes an audio signal containing speech as an input, and tries to produce the corresponding text representation \cite{huang2001spoken}. Figure \ref{fig:asr} presents the high-level block diagram of a typical speech recognition system, though other configurations are also possible. In practice, the actual implementation can in many cases contain more complicated connections between the different components, and the models can be packed together for optimization reasons \cite{yu2014automatic, kallasjoki2016}. The use of deep neural networks has also somewhat blurred the line between feature extraction and acoustic modeling \cite{hinton2012deep, kallasjoki2016}. \\

@@

However, $P(\bm{W} | \bm{O})$ is difficult to calculate directly \cite{gales2008application}, and therefore, needs to be transformed into a form that is easier to model statistically. The Bayes' theorem states that \cite[p.~10]{hori2013speech}:

@@

For finding the maximum, the denominator in equation \ref{eq:max2} can be discarded as it stays fixed, and equation \ref{eq:max} becomes:

@@

Here, $P(\bm{O} | \bm{W})$ is the probability of an acoustic observation sequence for a specific word sequence, corresponding to the acoustic model. $P(\bm{W})$ is the probability of a specific word or word sequence occurring, which is given directly by the language model. In practice, the probability product is often calculated as an addition in the logarithmic domain, with an additional weight $\alpha$ for controlling how much weight is given to the language model \cite[p.~200]{gales2008application}:

@@

The following subsections present a brief review of each individual part of the speech recognition process presented in figure \ref{fig:asr}. In other words, how solving equation \ref{eq:max3} is implemented in practice.

@@

	\includegraphics[width=\textwidth]{cepstrum.pdf}

@@

with $n$ ranging typically from two to four \cite[p.~210]{gales2008application}. The n-gram model has been the preferred way to construct language models for a long time \cite{gales2008application, kallasjoki2016, hirsimaki2009importance}. In practice, the n-gram probabilities estimated with equation \ref{eq:ml} need to be balanced, since n-grams that are present in the training material tend to get a too high probability and unseen sequences a too low probability. This process is called language model \textit{smoothing}, and can be achieved by distributing some probability mass from seen n-gram combinations to all unseen n-grams \cite{gales2008application}. \\\\

@@

In agglutinative languages like Finnish, Estonian, Hungarian and Turkish, words are formed primarily by concatenating suffixes to root words, as well as using compounding (i.e., joining words together) and inflections ("word bending") \cite{arisoy2008statistical, kurimo2006unlimited, hirsimaki2006unlimited}. This means that there are millions of regularly used word forms in these languages. Therefore, it is very hard to build a word-based vocabulary that would cover all the commonly used words \cite{arisoy2008statistical}. Though it has recently become possible to build n-gram models covering millions of words, reducing the vocabulary size is important for efficient models \cite{smit17boundaries}. For English, a 60K word lexicon can be sufficient for many tasks, whereas for Finnish, even a 500K word-based lexicon would not give the same level of performance \cite{arisoy2008statistical}. The problem is the large number of \textit{out-of-vocabulary} (OOV) words when dealing with limited vocabulary sizes. \textit{Sub-word modeling} has been successfully used to model agglutinative languages, reducing the size and complexity of the language model \cite{enarvi2017automatic, hirsimaki2006unlimited, hirsimaki2009importance, smit17boundaries}. In sub-word modeling, words are split into smaller units, which can be done based on grammatical rules or statistical techniques \cite{arisoy2008statistical}. A data-driven statistical method called \textit{Morfessor} is commonly used for morphological segmentation. The Morfessor uses unsupervised machine learning methods to find morpheme-like sub-word units called \textit{morphs}, based purely on raw text data \cite{arisoy2008statistical, kurimo2006unlimited}.

@@

In ASR systems, \textit{decoding} means computing the result based on the input and the statistical models, i.e., solving equation (\ref{eq:max3}), were the observation sequence $\bm{O}$ is the feature vector produced in feature extraction. The decoding process is fundamentally a search for the best matching word sequence over all possible word sequences, i.e., the \textit{search space} defined by the models \cite{pylkkonen2013towards}. However, implementing the search in practice is a non-trivial algorithmic challenge: The size of the search space is often huge, and grows exponentially with the number of words in an utterance \cite{pylkkonen2013towards}. Therefore, an exhaustive search over all possibilities is generally not feasible. Heuristic methods are required to limit the search space in some way. For this purpose, \textit{beam search} is commonly used \cite{pylkkonen2013towards, hori2013speech}. In beam search, only a limited number of most promising paths are kept while others are discarded in a process called \textit{pruning}. The \textit{Viterbi} dynamic programming algorithm is one such commonly used method \cite{huang2001spoken, yu2014automatic, hori2013speech}. \\\\
ASR systems are divided into \textit{online} and \textit{offline} decoders \cite{alumae2014full}. In online decoding, the input is coming continuously in real-time and speech recognition is performed concurrently with the input. In other words, speech recognition is performed in real-time and the results are typically displayed immediately. Conversely, in offline decoding the input is an existing (long) audio recording, which is then transcribed into text all at once. Offline decoding can be faster than real-time, meaning that the full transcription result is produced in less time than the length of the input file. For example, a ten minute audio file could be transcribed in five minutes. Online decoding sets some restrictions for the ASR system, with the primary constraint being the requirement of approximately real-time recognition. Therefore, a compromise between speed and accuracy is typically required in online recognition. For example, the size of the vocabulary may have to be limited for fast enough performance. \\\\
The use of \textit{weighted finite state transducers} (WFST) for representing the statistical models is popular in speech recognition, especially for LVCSR systems \cite{hori2013speech, mohri2008speech, smit17boundaries}. WFSTs can be understood as a finite automaton, consisting of a set of finite states and transitions between them. Each transition has an input label, output label and a weight for the transition. They are typically used to represent and store the acoustic model, language model and lexicon in one combined transducer, which offers many practical benefits. In the WFST--based speech recognition process, the acoustic model WFST transduces an acoustic state sequence into a phoneme sequence, the lexicon WFST transduces a phoneme sequence into a word sequence, and the language model WFST transduces a word sequence into a sentence. These transducers can be integrated into a single, large WFST that directly transduces an acoustic state sequence into a sentence. The power of weighted finite state transducers for ASR comes from that they can optimize the search space and remove redundancies. The end result is a highly-optimized static structure for fast decoding, enabling real-time decoding with very large vocabularies of over one million words even on an average personal computer. \cite[p.~4--6]{hori2013speech}

@@

Accurate speech recognition of everyday conversational speech has been the focus of much research in recent years, and remains challenging especially in noisy environments \cite{keronen2014approaching, kallasjoki2016, xiong2016achieving, li2014overview, keronen2010comparison, pylkkonen2013towards}. This particular area of ASR is the most relevant in the context of this work, as the Conversation Assistant is intended for use in everyday conversational situations. Conversational speech falls under the large vocabulary continuous speech recognition task, and is arguably the hardest area for online ASR. Conversational, or \textit{colloquial} speech, is typically quite informal and can differ noticeably from the standard, formal form of the language \cite{enarvi2017automatic}. Different dialects and pronunciations can pose a challenge, as the statistical models are only as good as the material they are trained with. The availability of suitable training data specifically for conversational speech can be problematic. Neural networks require a large amount of training data. Challenges in correct alignment and accurate transcripts. \\\\
Noise-robust speech recognition remains a challenge. Noise. Reverberation. \cite{kallasjoki2016, keronen2010comparison, keronen2014approaching, pohjalainen2009weighted, li2014overview}. Humans good at concentrating on one speaker among many others \cite{bronkhorst2000cocktail}.

@@

Acoustic model = speecon data: headset, lapel \& kaukomikit (1m), julkinen paikka ja auto. \cite[p.~43]{kallasjoki2016}. Kallasjoki, Remes, Keronen jne paperit. \\\\
Language model:  kielipankki, euroopan parlamentin käännetyt tekstit. \\\\
Morph-based vocabulary. \\\\
nnet2-online i-vector hr mfcc.

>>>>

\textit{Automatic speech recognition} (ASR) is the science and technology of transcribing spoken language into written words automatically using computers \cite{yu2014automatic, huang2001spoken}. The history of speech recognition research goes as far back as the 1950s, when the first steps were taken at the renowned Bell Laboratories \cite{gales2008application}. For a long time, working applications were comprised mostly of simple voice user interfaces and dictation systems with limited, application-specific vocabularies \cite{yu2014automatic, gales2008application}. The holy grail of speech recognition technology has arguably been real-time, speaker independent recognition of unlimited vocabularies, of which everyday conversational speech is a good example. In recent years, significant advances have been made especially in this area, referred to in ASR research as \textit{large vocabulary continuous speech recognition} (LVCSR) \cite{yu2014automatic, keronen2014approaching}. These advances have been enabled in large part thanks to the fast development and adoption of new machine learning techniques \cite{yu2014automatic, hinton2012deep}. In particular, \textit{deep neural networks} (DNN) with multiple hidden layers (hence the word "deep") have proven to be of great value for ASR systems \cite{yu2014automatic, hinton2012deep}. Machine learning itself has become widespread and practical owing to the advances in computer hardware and computational resources available, especially through the popularization of the so called \textit{general-purpose computing on graphics processing units} (GPGPU) technology for massively parallel computation, that has enabled efficient training of neural networks with relatively inexpensive and widely available computing hardware \cite{yu2014automatic, hinton2012deep}. \\\\
As a consequence, automatic speech recognition has gained a great deal of popularity during the last few years, and is starting to be widely used in practice in the commercial landscape and everyday technical applications, from smartphones and tablets to personal computers \cite{yu2014automatic}. Today, companies actively promoting their own speech recognition systems to average consumers include influential giants like Amazon, Apple, Google and Microsoft. Services based on speech recognition technology encompasses voice-controlled personal assistants, dictation, automatic captioning, voice-based search of audio and video content, and a wide variety of voice user interfaces \cite{yu2014automatic, li2014overview}. \\\\
While there are still some specific, and in many cases, quite substantial challenges for automatic speech recognition, in general it has become accurate and reliable enough to be used in many practical applications requiring continuous recognition of large vocabularies \cite{yu2014automatic, keronen2014approaching, mcgraw2016personalized}. This is demonstrated well by the speech recognition based services of the previously mentioned large technology companies, such as Apple's Siri voice assistant \cite{li2014overview}. At the same time, mobile smart devices (phones, tablets and laptops) and  wireless internet access have become ubiquitous in all developed countries of the world, offering a convenient platform for utilizing speech recognition technology in practically every conceivable place and situation \cite{yu2014automatic, mcgraw2016personalized}.

@@

In this work, the primary interest in automatic speech recognition is in using it in practice, instead of developing or improving upon some aspect of it. Therefore, this section focuses on providing a high-level, short overview of ASR systems. Likewise, detailed description of machine learning techniques and training methods fall outside this thesis. The focus in this thesis is on modern large vocabulary continuous speech recognition. \\\\ An automatic speech recognition system takes an audio signal containing speech as an input, and tries to produce the corresponding text representation \cite{huang2001spoken, li2014overview}. Figure \ref{fig:asr} presents the high-level block diagram of a typical speech recognition system, though other configurations are also possible. In practice, the actual implementation can in many cases contain more complicated connections between the different components, and the models can be packed together for optimization reasons \cite{yu2014automatic, kallasjoki2016}. The use of deep neural networks has also somewhat blurred the line between feature extraction and acoustic modeling \cite{hinton2012deep, kallasjoki2016}. \\

@@

However, $P(\bm{W} | \bm{O})$ is difficult to calculate directly \cite{gales2008application}, but it can be transformed into a form that is easier to model statistically. The Bayes' theorem states that \cite[p.~10]{hori2013speech}:

@@

For finding the maximum, the denominator in equation (\ref{eq:max2}) can be discarded as it stays constant, and equation (\ref{eq:max}) becomes:

@@

Here, $P(\bm{O} | \bm{W})$ is the probability of an acoustic observation sequence for a specific word sequence, corresponding to the acoustic model. $P(\bm{W})$ is the probability of a specific word or word sequence occurring, which is given by the language model. In practice, the probability product is often calculated as an addition in the logarithmic domain, with an additional weight $\alpha$ for controlling how much weight is given to the language model \cite[p.~200]{gales2008application}:

@@

The following subsections present a brief review of each individual part of the speech recognition process presented in figure \ref{fig:asr}. In other words, how solving equation (\ref{eq:max3}) is implemented in practice.

@@

	\includegraphics[trim={0.1cm 0cm 0cm 0cm}, clip, width=\textwidth]{cepstrum.pdf}

@@

with $n$ ranging typically from two to four \cite[p.~210]{gales2008application}. The n-gram model has been the preferred way to construct language models for a long time \cite{gales2008application, kallasjoki2016, hirsimaki2009importance}. In practice, the n-gram probabilities estimated with equation (\ref{eq:ml}) need to be balanced, since n-grams that are present in the training material tend to get a too high probability and unseen sequences a too low probability. This process is called language model \textit{smoothing}, and can be achieved by distributing some probability mass from seen n-gram combinations to all unseen n-grams \cite{gales2008application}. \\\\

@@

In agglutinative languages like Finnish, Estonian, Hungarian and Turkish, words are formed primarily by concatenating suffixes to root words, as well as using compounding (i.e., joining words together) and inflections ("word bending") \cite{arisoy2008statistical, kurimo2006unlimited, hirsimaki2006unlimited}. This means that there can be millions of regularly used word forms in these languages. Therefore, it is very hard to build a word-based vocabulary that would cover all the commonly used words \cite{arisoy2008statistical}. Though it has recently become possible to build n-gram models covering millions of words, reducing the vocabulary size is important for efficient models \cite{smit17boundaries}. For English, a 60K word lexicon can be sufficient for many tasks, whereas for Finnish, even a 500K word-based lexicon would not give the same level of performance \cite{arisoy2008statistical}. The problem is the large number of \textit{out-of-vocabulary} (OOV) words when dealing with limited vocabulary sizes. \textit{Sub-word modeling} has been successfully used to model agglutinative languages, reducing the size and complexity of the language model \cite{enarvi2017automatic, hirsimaki2006unlimited, hirsimaki2009importance, smit17boundaries}. In sub-word modeling, words are split into smaller units, which can be done based on grammatical rules or statistical techniques \cite{arisoy2008statistical}. A data-driven statistical method called the \textit{Morfessor} is commonly used for morphological segmentation. The Morfessor uses unsupervised machine learning methods to find morpheme-like statistical sub-word units called \textit{morphs}, based purely on raw text data \cite{arisoy2008statistical, kurimo2006unlimited}.

@@

In ASR systems, \textit{decoding} means computing the result based on the input and the statistical models, i.e., solving equation (\ref{eq:max3}), were the observation sequence $\bm{O}$ is the feature vector produced in feature extraction. The decoding process is fundamentally a search for the best matching word sequence over all possible word sequences, i.e., the \textit{search space} defined by the models \cite{pylkkonen2013towards}. However, implementing the search in practice requires efficient algorithmic solutions: The size of the search space is often huge, and grows exponentially with the number of words in an utterance \cite{pylkkonen2013towards}. Therefore, an exhaustive search over all possibilities is generally not feasible. Heuristic methods are required to limit the search space in some way. For this purpose, \textit{beam search} is commonly used \cite{pylkkonen2013towards, hori2013speech}. In beam search, only a limited number of most promising paths are kept while others are discarded in a process called \textit{pruning}. The \textit{Viterbi} dynamic programming algorithm is one such commonly used method \cite{huang2001spoken, yu2014automatic, hori2013speech}. \\\\
ASR systems are divided into \textit{online} and \textit{offline} decoders \cite{alumae2014full}. In online decoding, the input is coming continuously in real-time and speech recognition is performed concurrently with the input. In other words, speech recognition is performed in real-time and the results are typically displayed immediately. Conversely, in offline decoding the input is an existing (long) audio recording, which is then transcribed into text all at once. Offline decoding can be faster than real-time, meaning that the full transcription result is produced in less time than the length of the input file. For example, a ten minute audio file could be transcribed in five minutes. Online decoding sets some restrictions for the ASR system, with the primary constraint being the requirement of approximately real-time recognition. Therefore, a compromise between speed and accuracy is typically required in online recognition. For example, the size of the vocabulary may have to be limited for fast enough performance \cite{enarvi2017automatic, mcgraw2016personalized}. \\\\
The use of \textit{weighted finite state transducers} (WFST) for representing the statistical models is popular in speech recognition, especially for LVCSR systems \cite{hori2013speech, mohri2008speech, smit17boundaries}. WFSTs can be understood as a finite automaton, consisting of a set of finite states and transitions between them. Each transition has an input label, output label and a weight for the transition. They are typically used to represent and store the acoustic model, language model and lexicon in one combined transducer, which offers many practical benefits. In the WFST--based speech recognition process, the acoustic model WFST transduces an acoustic state sequence into a phoneme sequence, the lexicon WFST transduces a phoneme sequence into a word sequence, and the language model WFST transduces a word sequence into a sentence. These transducers can be integrated into a single, large WFST that directly transduces an acoustic state sequence into a sentence. The power of weighted finite state transducers for ASR comes from that they can optimize the search space and remove redundancies. The end result is a highly-optimized static structure for fast decoding, enabling real-time decoding with very large vocabularies of over one million words even on an average personal computer. \cite[p.~4--6]{hori2013speech}

@@

Accurate speech recognition of everyday conversational speech has been the focus of much research in recent years, and remains challenging especially in noisy environments \cite{keronen2014approaching, kallasjoki2016, xiong2016achieving, li2014overview, keronen2010comparison, pylkkonen2013towards}. This particular area of ASR is the most relevant in the context of this work, as the Conversation Assistant is intended for use in everyday conversational situations. Conversational speech falls under the large vocabulary continuous speech recognition task, and is arguably the hardest area for online ASR. Conversational, or \textit{colloquial} speech, is typically quite informal and can differ noticeably from the standard, formal form of the language \cite{enarvi2017automatic}. In Finnish, colloquial pronunciations are also written differently than the standard word form, due to the phonetic orthography \cite{enarvi2017automatic}. Consequently, colloquial speech causes more variations to language and increases the size of the vocabulary. For agglutinative languages, a vocabulary of many million words is required in order to cover all the spelling variations in conversational speech \cite{enarvi2017automatic}. Overall, these factors have made it significantly harder to achieve a good level of speech recognition performance for conversational speech, compared to most other tasks. \\\\
Different dialects and pronunciations can pose a challenge to ASR performance, as the statistical models are only as good as the material they are trained with. Neural networks require a large amount of training data, and sourcing suitable training data covering the wide variety of distinct speaking styles can be difficult \cite{mansikka17parliament}. In order to recognize colloquial variations such as heavy accents, tens or hundreds of hours of material is typically needed for good performance. In addition, the recordings have to be accurately transcribed and aligned correctly. Even if recording are available, transcriptions are typically not, and producing them is generally expensive and time consuming \cite{mansikka17parliament}.  \\\\
Conversations are often held in acoustically challenging situations with background noise and other competing speakers. For large-scale everyday usage of ASR, noise-robustness has become an integral part for good real-world recognition performance \cite{li2014overview}. More specialized ASR tasks can be performed in a quiet environment and with special, studio-grade microphone equipment, conforming to the limitations of the ASR system. For public usage, ASR systems have to cope with all possible environments and situations instead. Likewise, external microphone setups that could noticeably improve the SNR of the input signal are an unreasonable requirement for most everyday usage. Noise-robust speech recognition remains somewhat a challenge, though a large amount of research has been done to address this issue \cite{kallasjoki2016, keronen2010comparison, keronen2014approaching, li2014overview, qian2016very}. Fundamentally, the problem arises from distortions in the transmission path of the signal from the speaker to the microphone, with the effect that the observed feature sequence does not match the utterance of the speaker \cite{kallasjoki2016, li2014overview}. These distortions can be from sources such as traffic noise, reverberation, background music, and overlapping speech. \\\\
\textit{Noise-robust} speech recognition methods attempt to reduce the mismatch between the noisy observations and the models \cite{li2014overview, kallasjoki2016, keronen2010comparison}. One straightforward way to improve speech recognition in noisy conditions is to use \textit{multi-condition training}, where the acoustic model is trained with a combination of clean and noisy data. The feature extraction process can be improved further from the standard MFCC features to be more robust against noise. Additionally, \textit{feature enhancement} methods have been presented that adaptively compensate for distortions in the features, such as using imputation methods for reconstructing missing data. Besides training noise robust models, the model parameters can adapted in real-time to the prevailing conditions \cite{kallasjoki2016}.

@@

Acoustic model = speecon data: headset, lapel \& kaukomikit (1m), julkinen paikka ja auto. \cite[p.~43]{kallasjoki2016}. Kallasjoki, Remes, Keronen jne paperit. \\\\
Language model:  kielipankki, euroopan parlamentin käännetyt tekstit. \\\\
Morph-based vocabulary. \\\\
nnet2-online i-vector hr mfcc. \\\\
i-vector speaker adaptation \cite{xiong2016achieving, kaldi}.
